QPLIB: A Library of Quadratic Programming Instances

### 2018-02-04 QPLIB-2.2.tex
consistent title case in section headers

% -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -
\subsubsection{Examples and reformulations}\label{ssec:reform}

@@

Another relevant reformulation trick concerns the fact that, as soon as quadratic constraints are allowed, then there is no loss of generality in assuming a linear objective function. Indeed, any \textit{Q**} (\textit{C*C}) problem can always be rewritten as
%
\begin{align*}
 \min \;\;
 & x^0 \\
 %
 & \hspace*{-5mm} -\infty \leq {\textstyle \half} x^\top Q^0 x + b^0 x \leq x^0 & \\
 %
 & c^i_l \leq {\textstyle \half} x^\top Q^i x + b^i x \leq c^i_u & i \in \Mcm \\
 %
 & l_j \leq x_j \leq u_j & j \in \Mcn  \\
 %
 & x_j \in \mathbb{Z} & j \in \Mcz
\end{align*}
%
i.e., a \textit{L*Q} (\textit{L*C}) pronlem. Hence, it is clear that quadratic constraints are, in a well-defined sense, the most general situation (cf.~also the result above about hardness of \textit{LIQ}).

@@

% -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -
\subsubsection{QP classes}\label{ssec:classes}

@@

 \item \emph{Convex Continuous Quadratic Programs} \textit{CCC} can be solved in polynomial time by Interior-Point techniques \cite{Wright97}; the simpler \textit{CCL} can also be solved by means of ``simplex-like'' techniques, usually referred to as active-set methods \cite{Dost09}. Actually, a slightly larger class of problems can be solved with Interior-Point methods: those that can be represented by Second-Order Cone Programs. When written as QPs the corresponding $Q^i$ may not be positive semidefinite, but nonetheless such problems can be efficiently solved. Of course just as for \textit{LCL}, these problems may still require considerable computational effort when the size of the instance grows. In this sense, like in the linear case, there is a significant distinction between solvers that need all the data of QP to work, and those that are ``matrix-free'', i.e., only require the application of simple operations (typically, matrix-vector products) with the problem data. While when building our instance library we never exploited such characteristics, since they are not amenable to standard modelling tools, but this may be relevant for the solution of very-large-scale \textit{CIC}.

>>>>

% -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -
\subsubsection{Examples and Reformulations}\label{ssec:reform}

@@

Another relevant reformulation trick concerns the fact that, as soon as quadratic constraints are allowed, then there is no loss of generality in assuming a linear objective function. Indeed, any \textit{Q**} (\textit{C*C}) problem can always be rewritten as
%
\begin{align*}
 \min \;\;
 & x^0 \\
 %
 & \hspace*{-5mm} -\infty \leq {\textstyle \half} x^\top Q^0 x + b^0 x \leq x^0 & \\
 %
 & c^i_l \leq {\textstyle \half} x^\top Q^i x + b^i x \leq c^i_u & i \in \Mcm \\
 %
 & l_j \leq x_j \leq u_j & j \in \Mcn  \\
 %
 & x_j \in \mathbb{Z} & j \in \Mcz
\end{align*}
%
i.e., a \textit{L*Q} (\textit{L*C}) problem. Hence, it is clear that quadratic constraints are, in a well-defined sense, the most general situation (cf.~also the result above about hardness of \textit{LIQ}).

@@

% -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -  -
\subsubsection{QP Classes}\label{ssec:classes}

@@

 \item \emph{Convex Continuous Quadratic Programs} \textit{CCC} can be solved in polynomial time by Interior-Point techniques \cite{Wright97}; the simpler \textit{CCL} can also be solved by means of ``simplex-like'' techniques, usually referred to as active-set methods \cite{Dost09}. Actually, a slightly larger class of problems can be solved with Interior-Point methods: those that can be represented by Second-Order Cone Programs. When written as QPs the corresponding $Q^i$ may not be positive semidefinite, but nonetheless such problems can be efficiently solved. Of course just as for \textit{LCL}, these problems may still require considerable computational effort when the size of the instance grows. In this sense, like in the linear case, there is a significant distinction between solvers that need all the data of QP to work, and those that are ``matrix-free'', i.e., only require the application of simple operations (typically, matrix-vector products) with the problem data. While when building our instance library we never exploited such characteristics, since they are not amenable to standard modeling tools, but this may be relevant for the solution of very-large-scale \textit{CIC}.

### 2018-02-04 QPLIB-3.tex
consistent title case in section headers

In this section we describe the procedure we adopted to gather the
instances. In January $2014$, we issued an online call for instances
using the main international mailing lists of the mathematical
optimization and numerical analysis communities, reaching in this way
the largest possible set of interested researchers and practitioners.
The call remained open for 10 months, during which we received a large
number of contributions of different nature. The instances we gathered
come both from theoretical studies as well as from real-world
applications.

In addition to spontaneous contribution we analysed the other generic
libraries of instances available  on internet and containing QP
instances. Namely, the libraries from which we gathered instances are
%
\begin{itemize}
 \item the \texttt{BARON} library
 \url{http://www.minlp.com/nlp-and-minlp-test-problems};
%
\item the \texttt{CUTEst} library
 \url{https://ccpforge.cse.rl.ac.uk/gf/project/cutest};
%
\item the \texttt{GAMS Performance} libraries
 \url{http://www.gamsworld.org/performance/performlib.htm};
%
\item the \texttt{MacMINLP} library
 \url{https://wiki.mcs.anl.gov/leyffer/index.php/MacMINLP};
%
\item the \texttt{Meszaros} library
 \url{http://www.doc.ic.ac.uk/~im/00README.QP};
%
\item the \texttt{MINLP} library
 \url{http://www.gamsworld.org/minlp/minlplib.htm};
%
\item the \texttt{POLIP} library
 \url{http://polip.zib.de/pipformat.php}.
\end{itemize}

Other quadratic instances were found in online libraries devoted to
specific QP problems as Max-Cut, Quadratic Assignment, Portfolio
Optimization, and several others. \textcolor{red}{In addition, we mention
that other generic libraries exist, e.g., 
Conic library CBLIB (\url{http://cblib.zib.de}) and 
MIPLIB 2010 (\url{http://miplib.zib.de/}), to mention just a few.}

@@

The number of positive (i.e., larger than $10^{-12}$) and negative (i.e., smaller than $-10^{-12}$) eigenvalues of $Q^0$ allowed us to identify the
objective function type, as in presence of at least one negative (positive) eigenvalue
the objective function is nonconvex (nonconcave). Finally, as far as the constraint types
are concerned, we collected the following information:
%
\begin{itemize}
 \item the number of linear constraints, % (\emph{\# lin}),
 %
 \item the number of quadratic constraints, % (\emph{\# quad}),
 %
 \item the number of convex constraints, and % (\emph{\# conv}),
 %
 \item the number of variable bounds (for non-binary variables). %(\emph{\# box}).
\end{itemize}
%
A constraint is considered quadratic if it contains at least one nonzero in
a quadratic term (if present). Among the quadratic constraints, the ones whose
Hessians have only non-negative eigenvalues (when $c_u^i < \infty$) and
and non-positive eigenvalues (when $c_l^i > - \infty$)
are classified as convex constraints; thus, a quadratic constraint with
two sided, finite bounds is non-convex.
Note that this might occasionally lead us to classify some instances that have conic constraints as non-convex ones, although their feasible region is in fact convex---fortunately, only some solvers are capable of properly exploiting this property.
 All this information allowed us to analyse the gathered instances and to perform the
filters described in the the next paragraph.

@@

During the development of the library, a discussion ensued about
the expected goals that we wished to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item to represent as far as possible all the different categories of QP
       problems;
 %
 \item to gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item to include, for each of the categories, a set of well-diversified
       instances; and
 %
 \item to obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve such goals, we performed the following two filters, applied in
a cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S\ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in GAMS, which number depends on the category
       of the instance under consideration, cf.~Table \ref{t:solvers}.
       We then filtered according to a first
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analysed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives (e.g., very similar size, same donor,\ldots). Finally, in order to only
       keep computationally challenging instances we ran a a complete solver for QGQ with a time limit of 120 seconds; all the
       instances which have been solved to proven optimality within this time limit
       were discarded.
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $319$ discrete instances and
$134$ continuous instances.

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Analysis of the final set of instances}\label{subsec:final set}

\textcolor{red}{We now analyse the features of the instances selected to be part of the
 library. In Table \ref{tab:DDDDD}, we provide a global overview. The instances have been divided in \textit{continuous} vs \textit{discrete} and \textit{convex} vs \textit{non-convex}, forming in this way, a  classification of 4 macro categories. As previously mentioned, an instance is classified \textit{discrete} if it contains at least a binary or integer variable, and \textit{continuous} otherwise. On the other hand, an instance is classified as \textit{non-convex} if the objective function is non-convex and/or at least one of the constraints is  non-convex, and \textit{convex} otherwise. }

@@

On the theme of nonconvexity, Figure~\ref{fig:pic_neg_eig} (left) focuses on
the instances with a quadratic objective function, ordered by 
 percentage of ``problematic'' \textcolor{red}{(defined using a tollerance of XXX)} \todo{please define the tolerance used to identify the ``Problematic'' eigenvalues and change the label of the plot}\ eigenvalues in the Hessian $Q^0$ (vertical axis), by which
we mean negative eigenvalues in case of a minimization problem and positive eigenvalues in case of a maximization problem.
The instances are mostly clustered around two values. About 25\% of the instances have a convex (if minimization) or concave (if maximization) objective function,
i.e., they have 0\% of ``problematic'' eigenvalues. Among the others, a vast
majority has around 50\% of ``problematic'' eigenvalues. However, instances
with high or low percentages of ``problematic'' eigenvalues are present, too.

>>>>

In this section we describe the procedure we adopted to gather the
instances. In January $2014$, we issued an online call for instances
using the main international mailing lists of the mathematical
optimization and numerical analysis communities, reaching in this way
the largest\todo{AG: a large?} possible set of interested researchers and practitioners.
The call remained open for ten months, during which we received a large
number of contributions of different nature. The instances we gathered
come both from theoretical studies as well as from real-world
applications.

In addition to spontaneous contribution we analyzed the other generic
libraries of instances available  on the internet and containing QP
instances. Namely, the libraries from which we gathered instances are
%
\begin{itemize}
 \item the \texttt{BARON} library
 \url{http://www.minlp.com/nlp-and-minlp-test-problems};
%
\item the \texttt{CUTEst} library
 \url{https://ccpforge.cse.rl.ac.uk/gf/project/cutest};
%
\item the \texttt{GAMS Performance} libraries
 \url{http://www.gamsworld.org/performance/performlib.htm};
%
\item the \texttt{MacMINLP} library
 \url{https://wiki.mcs.anl.gov/leyffer/index.php/MacMINLP};
%
\item the \texttt{Meszaros} library
 \url{http://www.doc.ic.ac.uk/~im/00README.QP};
%
\item the \texttt{MINLP} library
 \url{http://www.gamsworld.org/minlp/minlplib.htm};
%
\item the \texttt{POLIP} library
 \url{http://polip.zib.de/pipformat.php}.
\end{itemize}

Other quadratic instances were found in online libraries devoted to
specific QP problems as Max-Cut, Quadratic Assignment, Portfolio
Optimization, and several others. \textcolor{red}{In addition, we mention
that other generic libraries exist, e.g., 
Conic library CBLIB (\url{http://cblib.zib.de}) and 
MIPLIB~2010 (\url{http://miplib.zib.de/}), to mention just a few.}\todo{AG: maybe cite CBLIB paper \cite{Friberg2016}}

@@

The number of positive (i.e., larger than $10^{-12}$) and negative (i.e., smaller than $-10^{-12}$) eigenvalues of $Q^0$ allowed us to identify the
objective function type, as in presence of at least one negative (positive) eigenvalue
the objective function is nonconvex (nonconcave). Finally, as far as the constraint types
are concerned, we collected the following information:
%
\begin{itemize}
 \item the number of linear constraints, % (\emph{\# lin}),
 %
 \item the number of quadratic constraints, % (\emph{\# quad}),
 %
 \item the number of convex constraints, and % (\emph{\# conv}),
 %
 \item the number of variable bounds (for non-binary variables). %(\emph{\# box}).
\end{itemize}
%
A constraint is considered quadratic if it contains at least one nonzero in
a quadratic term (if present). Among the quadratic constraints, the ones whose
Hessians have only non-negative eigenvalues (when $c_u^i < \infty$) and
and non-positive eigenvalues (when $c_l^i > - \infty$)
are classified as convex constraints; thus, a quadratic constraint with
two sided, finite bounds is non-convex.
Note that this might occasionally lead us to classify some instances that have conic constraints as non-convex ones, although their feasible region is in fact convex---fortunately, only some solvers are capable of properly exploiting this property.
 All this information allowed us to analyze the gathered instances and to perform the
filters described in the the next paragraph.

@@

During the development of the library, a discussion ensued about
the expected goals that we wished to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item to represent as far as possible all the different categories of QP
       problems;
 %
 \item to gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item to include, for each of the categories, a set of well-diversified
       instances; and
 %
 \item to obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve such goals, we performed the following two filters, applied in
a cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S\ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in GAMS, which number depends on the category
       of the instance under consideration, cf.~Table \ref{t:solvers}.
       We then filtered according to a first
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analyzed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives (e.g., very similar size, same donor,\ldots). Finally, in order to only
       keep computationally challenging instances we ran a a complete solver for QGQ with a time limit of 120 seconds; all the
       instances which have been solved to proven optimality within this time limit
       were discarded.
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $319$ discrete instances and
$134$ continuous instances.

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{Analysis of the Final Set of Instances}\label{subsec:final set}

\textcolor{red}{We now analyze the features of the instances selected to be part of the
 library. In Table \ref{tab:DDDDD}, we provide a global overview. The instances have been divided in \textit{continuous} vs \textit{discrete} and \textit{convex} vs \textit{non-convex}, forming in this way, a  classification of 4 macro categories. As previously mentioned, an instance is classified \textit{discrete} if it contains at least a binary or integer variable, and \textit{continuous} otherwise. On the other hand, an instance is classified as \textit{non-convex} if the objective function is non-convex and/or at least one of the constraints is  non-convex, and \textit{convex} otherwise. }

@@

On the theme of nonconvexity, Figure~\ref{fig:pic_neg_eig} (left) focuses on
the instances with a quadratic objective function, ordered by 
 percentage of ``problematic'' \textcolor{red}{(defined using a tolerance of XXX)} \todo{please define the tolerance used to identify the ``Problematic'' eigenvalues and change the label of the plot}\ eigenvalues in the Hessian $Q^0$ (vertical axis), by which
we mean negative eigenvalues in case of a minimization problem and positive eigenvalues in case of a maximization problem.
The instances are mostly clustered around two values. About 25\% of the instances have a convex (if minimization) or concave (if maximization) objective function,
i.e., they have 0\% of ``problematic'' eigenvalues. Among the others, a vast
majority has around 50\% of ``problematic'' eigenvalues. However, instances
with high or low percentages of ``problematic'' eigenvalues are present, too.
